Inferring the existence and direction of causal associations in the face of measurement error

### 2017-01-31 docs/manuscript.Rmd
updated

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
	display=FALSE
)

>>>>

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables. The causal inference test attempts to delineate between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
	display=FALSE
)

### 2017-01-31 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 January 2017}

@@

A common approach for inferring causal relationships between two traits
is to use genetic instruments in mediation-based analyses. Here we show
that when measurement error is present, commonly-used tools predicated
on this approach are susceptible to making erroneous inference of both
the existence of causality and the causal direction, and that increasing
sample size, rather than reducing bias, can often have the effect of
increasing certainty in the wrong answer. We demonstrate that Mendelian
randomization (MR) is a technique for causal inference that is robust to
measurement error, and here we extend the method with new sensitivity
analyses to estimate the direction of causality between correlated
variables even when the biology of the instrument is unknown. This
extension is less susceptible to measurement error than mediation-based
analyses. We use it to demonstrate empirically that inferring the
direction of causality between correleted DNA methylation and gene
expression levels is highly susceptible to differences in measurement
error between the two assaying platforms. Other 'omic platforms are
likely to be similarly susceptible, necessitating the use of sensitivity
analyses to evaluate reliability.

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding\textsuperscript{1--3}. In response to this, genetic
instrumentation has emerged as a technique for improving the reliability
of causal inference in observational data, and with the coincident rise
in genome-wide association studies it is now a prominent tool that is
applied in several different guises\textsuperscript{3--6}. However,
potential pitfalls still persist and one that is often neglected is the
influence of non-differential measurement error on the reliability of
causal inference.

Non-differential measurement error (or observational error, henceforth
referred to simply as measurement error) is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{7,8}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{9,10}. Relatedly, if the measurement of biological
interest is the methylation level in a T cell, then measurement error of
this value can be introduced by using methylation levels from whole
blood samples because the measured value will be an assay of many cell
types\textsuperscript{11}.

Measurement error will of course arise in more basic data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}.

Given the near-ubiquitous presence of measurement error in phenomic data
it is vital to understand its impact on the tools we use for causal
inference. An established study design that can provide information
about causality is randomisation. Given the hypothesis that trait A
(henceforth referred to as the exposure) is causally related to trait B
(henceforth referred to as the outcome), randomisation can be employed
to assess the causal nature of the association by randomly splitting the
sample into two groups, subjecting one group to the exposure and leaving
the other as a control. The association between the exposure and the
outcome in this setting provides a robust estimate of the causal
relationship. This provides the theoretical basis behind randomised
control trials, but in practice randomisation is often impossible to
implement in an experimental context due to cost, scale or inability to
manipulate the exposure. The principle, however, can be employed in
extant observational data through the use of genetic variants associated
with the exposure (instruments), where the inheritance of an allele
serves as a random lifetime allocation of differential exposure
levels\textsuperscript{15,16}. Two statistical approaches to exploiting
the properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{27 January 2017}

@@

Genetic associations have the desirable property of ensuring the causal
direction is from genotype to trait. Mediation based approaches such as
the causal inference test (CIT) use this property to infer the causal
direction between two traits, while Mendelian randomization uses the
property to estimate the causal effect size if knowledge of the

by evaluating if the genetic association on the consequential trait is
abbrogated by adjusting for the causal trait. Mendelian randomization,
by contrast assumes a priori knowledge of which of the two traits the
genetic variant is acting on primarily, and then creates a proxy to
estimate its causal effect on

A common approach for inferring causal relationships between two traits
is to use genetic instruments in mediation-based analyses. Here we show
that when measurement error is present, widely used tools predicated on
this approach, such as the causal inference test (CIT) are susceptible
to making erroneous inference of both the existence of causality and the
causal direction. Increasing sample size for these tests, rather than
reducing bias, can often have the effect of increasing certainty in the
wrong answer. We demonstrate that Mendelian randomization (MR) is a
technique for causal inference that is relatively robust to measurement
error, and here we extend the method with new sensitivity analyses to
estimate the direction of causality between correlated variables even
when the biology of the instrument is unknown. This extension is less
susceptible to measurement error than mediation-based analyses and only
requires GWAS summary data. We use it to demonstrate empirically that
inferring the direction of causality between correleted DNA methylation
and gene expression levels is highly susceptible to differences in
measurement error between the two assaying platforms. Other 'omic
platforms are likely to be similarly susceptible, necessitating the use
of sensitivity analyses to evaluate reliability.

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding\textsuperscript{1--3}. In response to this, genetic
instrumentation has emerged as a technique for improving the reliability
of causal inference in observational data, and with the coincident rise
in genome-wide association studies it is now a prominent tool that is
applied in several different guises\textsuperscript{3--6}. However,
potential pitfalls remain and one that is often neglected is the
influence of non-differential measurement error on the reliability of
causal inference.

Measurement error (or observational error, henceforth referred to simply
as measurement error) is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to
avoid\textsuperscript{7,8}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but comes with the
problem of having imperfect resolution, for instance methylation levels
as measured by the Illumina450k chip are prone to have some amount of
noise around the true value due to imperfect
sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{11}.

Measurement error will of course arise in more basic data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality
is randomisation. Given the hypothesis that trait A (henceforth referred
to as the exposure) is causally related to trait B (henceforth referred
to as the outcome), randomisation can be employed to assess the causal
nature of the association by randomly splitting the sample into two
groups, subjecting one group to the exposure and leaving the other as a
control. The association between the exposure and the outcome in this
setting provides a robust estimate of the causal relationship. This
provides the theoretical basis behind randomised control trials, but in
practice randomisation is often impossible to implement in an
experimental context due to cost, scale or inability to manipulate the
exposure. The principle, however, can be employed in extant
observational data through the use of genetic variants associated with
the exposure (instruments), where the inheritance of an allele serves as
a random lifetime allocation of differential exposure
levels\textsuperscript{15,16}. Two statistical approaches to exploiting
the properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).
